---
title: "Analysis of exposures with a neutral discourse context"
output: html_notebook
---

This notebook analyzes the generalization patterns of novel dative-verb learners based on the factors of exposure contexts which situate the novel verb in a neutral discourse context where only the agent is given. For example:

```
DO: mommy [verbed] me a ball.
PP: mommy [verbed] a ball to me .
```

```{r, echo=FALSE,message=FALSE}
library(tidyverse)
library(fs)
library(jsonlite)
library(lme4)
library(lmerTest)
library(glue)
library(ggtext)
library(emmeans)
library(broom.mixed)
library(DT)

theme_set(
  theme_bw(base_size = 17, base_family = "Times") +
    theme(
      legend.position = "top",
      axis.text = element_text(color="black")
    )
)
```

## Loading Data (code hidden)
```{r, echo=FALSE,message=FALSE}
# loading data
pronouns = c("him", "her", "them", "me", "us", "it", "something", "someone")
propernouns = c("bear", "bear over there", "mommy", 
                "daddy", "grandma", "cat", "dog", "dog", 
                "bert", "elmo", "dog outside the house", 
                "cat outside the house", "teddy", "dolly",
                "cookies", "cheerios")

propro = c(pronouns, propernouns)

mode = "_valtest_vbd_no_discourse"

adaptation1 <- stream_in(file(glue("../data/experiments/single_stimuli_dative_simulation{mode}/adaptation.jsonl"))) %>%
  as_tibble() %>%
  mutate(
    canonicality = case_when(
      theme_markedness == recipient_markedness ~ "unmarked",
      theme_pronominality == "pronoun" & recipient_pronominality == "pronoun" ~ "unmarked",
      theme_definiteness == "definite" & theme_pronominality != "pronoun" & recipient_pronominality == "pronoun" ~ "nc-marked",
      theme_markedness == "unmarked" & recipient_markedness == "marked" ~ "nc-marked",
      theme_markedness == "marked" & recipient_markedness == "unmarked" ~ "marked"
    )
  )

adaptation2 <- stream_in(file(glue("../data/experiments/single_stimuli_dative_simulation{mode}2/adaptation.jsonl"))) %>%
  as_tibble() %>%
  mutate(
    canonicality = case_when(
      theme_markedness == recipient_markedness ~ "unmarked",
      theme_pronominality == "pronoun" & recipient_pronominality == "pronoun" ~ "unmarked",
      theme_definiteness == "definite" & theme_pronominality != "pronoun" & recipient_pronominality == "pronoun" ~ "nc-marked",
      theme_markedness == "unmarked" & recipient_markedness == "marked" ~ "nc-marked",
      theme_markedness == "marked" & recipient_markedness == "unmarked" ~ "marked"
    )
  )

adaptation3 <- stream_in(file(glue("../data/experiments/single_stimuli_dative_simulation{mode}3/adaptation.jsonl"))) %>%
  as_tibble() %>%
  mutate(
    canonicality = case_when(
      theme_markedness == recipient_markedness ~ "unmarked",
      theme_pronominality == "pronoun" & recipient_pronominality == "pronoun" ~ "unmarked",
      theme_definiteness == "definite" & theme_pronominality != "pronoun" & recipient_pronominality == "pronoun" ~ "nc-marked",
      theme_markedness == "unmarked" & recipient_markedness == "marked" ~ "nc-marked",
      theme_markedness == "marked" & recipient_markedness == "unmarked" ~ "marked"
    )
  )

td_ids = adaptation3 %>% pull(item)
# td_ids = c()

adaptation <- bind_rows(
  bind_rows(adaptation1, adaptation2) %>%
    filter(!item %in% td_ids),
  adaptation3
) %>%
  mutate(
    distinctiveness = case_when(
      theme %in% propro & !recipient %in% propro ~ "distinct",
      theme %in% propro & recipient %in% propro ~ "not_distinct",
      !theme %in% propro & recipient %in% propro ~ "distinct",
      !theme %in% propro & !recipient %in% propro ~ "not_distinct"
    )
  )

generalization <- stream_in(file(glue("../data/experiments/single_stimuli_dative_simulation{mode}/generalization.jsonl"))) %>%
  as_tibble()


feature_configs <- adaptation %>% 
  distinct(theme_pronominality, theme_animacy, theme_length, theme_definiteness, theme_markedness, recipient_pronominality, recipient_animacy, recipient_length, recipient_definiteness, recipient_markedness, canonicality) %>%
  rowwise() %>%
  mutate(
    feature_config = glue::glue_collapse(str_sub(c(theme_pronominality, theme_animacy, theme_length, theme_definiteness, theme_markedness, recipient_pronominality, recipient_animacy, recipient_length, recipient_definiteness, recipient_markedness), 1, 1))
  )


results <- bind_rows(
  bind_rows(
    dir_ls(glue("../data/results/single_stimuli_dative_simulation{mode}/"), recurse = TRUE, regexp = "*lr_results_hypwise.csv") %>%
      map_df(read_csv, .id = "model") %>%
      mutate(
        model = str_extract(model, glue("(?<=simulation{mode}/)(.*)(?=/best_lr_results_hypwise.csv)"))
      ),
    dir_ls(glue("../data/results/single_stimuli_dative_simulation{mode}2/"), recurse = TRUE, regexp = "*lr_results_hypwise.csv") %>%
      map_df(read_csv, .id = "model") %>%
      mutate(
        model = str_extract(model, glue("(?<=simulation{mode}2/)(.*)(?=/best_lr_results_hypwise.csv)"))
      ) 
  ) %>%
    filter(!item_id %in% td_ids),
  dir_ls(glue("../data/results/single_stimuli_dative_simulation{mode}3/"), recurse = TRUE, regexp = "*lr_results_hypwise.csv") %>%
    map_df(read_csv, .id = "model") %>%
    mutate(
      model = str_extract(model, glue("(?<=simulation{mode}3/)(.*)(?=/best_lr_results_hypwise.csv)"))
    ),
) %>%
  mutate(
    seed = as.numeric(str_remove(model, "smolm-autoreg-bpe-seed_"))
  )

results <- results %>%
  inner_join(feature_configs %>% rename(adaptation_feature_config = feature_config)) %>%
  inner_join(adaptation %>% rename(adaptation_dative = dative)) %>%
  # filter(seed %in% c(6, 28, 221, 1024, 1102, 1729))
  filter(seed %in% c(6, 28, 221, 394, 496, 1024, 1102, 1729, 2309, 8128))

best_results <- results %>%
  pivot_wider(names_from = state, values_from = logprob) %>%
  mutate(
    diff = best - initial
  )
```

## Cross Dative Generalization

Investigating alternate-form generalization patterns of the learner when it is exposed to the double object dative vs. prepositional object dative.

```{r, echo=FALSE}
results %>%
  pivot_wider(names_from = state, values_from = logprob) %>%
  mutate(diff = best - initial) %>%
  # inner_join(adaptation %>% rename(adaptation_dative = dative)) %>%
  # filter(canonicality == "marked") %>%
  group_by(adaptation_dative, generalization_dative) %>%
  summarize(
    n = n(),
    # ste = 1.96 * plotrix::std.error((best-initial)/initial),
    # logprob = mean((best-initial)/initial),
    ste = 1.96 * plotrix::std.error(best),
    logprob = mean(best)
  ) %>%
  ungroup() %>%
  filter(adaptation_dative != generalization_dative) %>%
  mutate(
    adaptation_dative = str_to_upper(adaptation_dative),
    generalization_dative = str_to_upper(generalization_dative),
    experiment = glue::glue("{adaptation_dative} --> {generalization_dative}")
  ) %>%
  ggplot(aes(experiment, logprob)) +
  # geom_boxplot()
  # geom_jitter(size = 2, width = 0.1)
  geom_point(size = 3, color = "steelblue") +
  geom_linerange(aes(ymin = logprob-ste, ymax=logprob+ste), color = "steelblue") +
  # facet_wrap(~model, scales="free") +
  labs(
    x = "Generalization Experiment",
    y = "Average Log P(construction)"
  ) +
  theme(
    axis.text.x = element_markdown(),
    panel.grid = element_blank()
  )
```

## Conwell's "me" finding (2019):

Pronominal recipients enable DO comprehension in children, so if we were to believe that comprehension is a precondition to generalization, then pronominal recipients should also encourage generalization:

```{r, echo=FALSE}
best_results %>%
  filter(adaptation_dative == "do") %>%
  mutate(
    config2 = recipient_pronominality,
    config1 = ""
  ) %>%
  group_by(config1, config2, adaptation_dative, generalization_dative) %>%
  summarize(
    n = n(),
    ste = 1.96 * plotrix::std.error(best),
    logprob = mean(best)
  ) %>%
  ungroup() %>%
  mutate(
    adaptation_dative = str_to_upper(adaptation_dative),
    generalization_dative = str_to_upper(generalization_dative),
    exposure = glue::glue("Exposure = {str_to_upper(adaptation_dative)}")
  ) %>%
  filter(adaptation_dative != generalization_dative) %>%
  ggplot(aes(config2, logprob, color = generalization_dative)) +
  geom_point(size = 3) +
  geom_linerange(aes(ymin = logprob-ste, ymax=logprob+ste)) +
  # facet_grid(config1~exposure, scales = "free") +
  facet_wrap(~exposure, scales = "free") +
  scale_color_manual(values = c("#00BFC4")) +
  theme(
    panel.grid = element_blank()
  ) +
  labs(
    x = "Feature Configuration",
    y = "log P(alt-form)",
    color = "Generalization Dative"
  )
```


## Replicating Arunachalam (2017):

Only looking at definite themes and recipients:

```{r, echo=FALSE, fig.width=10}
best_results %>%
  filter(theme_definiteness == "definite", recipient_definiteness == "definite") %>%
  filter(recipient_animacy == "animate") %>%
  mutate(
    # config2 = glue("{recipient_animacy}-R\n{theme_animacy}-T"),
    config2 = glue("{theme_animacy}-theme"),
    config1 = ""
  ) %>%
  group_by(config1, config2, adaptation_dative, generalization_dative) %>%
  summarize(
    n = n(),
    ste = 1.96 * plotrix::std.error(best),
    logprob = mean(best)
  ) %>%
  ungroup() %>%
  mutate(
    adaptation_dative = str_to_upper(adaptation_dative),
    generalization_dative = str_to_upper(generalization_dative),
    exposure = glue::glue("Exposure = {str_to_upper(adaptation_dative)}")
  ) %>%
  filter(adaptation_dative != generalization_dative) %>%
  ggplot(aes(config2, logprob, color = generalization_dative)) +
  geom_point(size = 3) +
  geom_linerange(aes(ymin = logprob-ste, ymax=logprob+ste)) +
  # facet_grid(config1~exposure, scales = "free") +
  facet_wrap(~exposure, scales = "free")+
  theme(
    panel.grid = element_blank()
  ) +
  labs(
    x = "Feature Configuration",
    y = "log P(alt-form)",
    color = "Generalization Dative"
  )
```



```{r, echo=FALSE}
coded_results <- best_results %>%
  # filter(!str_detect(sentence, "(dolly|teddy)")) %>%
  # filter(adaptation_dative == "do", generalization_dative == "pp") %>%
  mutate(
    # recipient_pronominality = factor(recipient_pronominality),
    # theme_pronominality = factor(theme_pronominality),
    recipient_pronominality = case_when(
      recipient_pronominality == "pronoun" ~ 0.5,
      TRUE ~ -0.5
    ),
    theme_pronominality = case_when(
      theme_pronominality == "pronoun" ~ 0.5,
      TRUE ~ -0.5
    ),
    recipient_animacy = case_when(
      recipient_animacy == "animate" ~ 0.5,
      TRUE ~ -0.5
    ),
    theme_animacy = case_when(
      theme_animacy == "animate" ~ 0.5,
      TRUE ~ -0.5
    ),
    theme_definiteness = case_when(
      theme_definiteness == "definite" ~ 1,
      TRUE ~ 0
    ),
    recipient_definiteness = case_when(
      recipient_definiteness == "definite" ~ 1,
      TRUE ~ 0
    ),
    theme_length = case_when(
      theme_length == "short" ~ 1,
      TRUE ~ 0
    ),
    recipient_length = case_when(
      recipient_length == "short" ~ 1,
      TRUE ~ 0
    ),
    distinctiveness = case_when(
      distinctiveness == "distinct" ~ 1,
      TRUE ~ 0
    ),
    model = factor(model),
    item_id = factor(item_id),
    hypothesis_id = factor(hypothesis_id)
  )
```

## Interaction terms

The design of interaction terms is a bit tricky. Due to our design, the pronominality might interact with the animacy of a given argument, but animacy and pronominality might also interact across arguments -- e.g., if theme is inanimate then LMs might prefer animate recipients (if they are sensitive to the semantics of the dative construction -- caused possession/motion). This means we should include the following interaction terms:

```
- theme_animacy * theme_pronominality
- recipient_animacy * recipient_pronominality
- theme_animacy * recipient_animacy
- theme_pronominality * recipient pronominality
```


## DO --> PP analysis

```{r}
do_pp <- coded_results %>%
  filter(adaptation_dative == "do", generalization_dative == "pp")

# interactions between theme-animacy and recipient animacy; theme-pronominality and recipient-pronominality

fit_tara_tprp_interaction_dopp <- lmer(best ~ 
                                    theme_animacy + theme_pronominality +
                                    recipient_animacy + recipient_pronominality +
                                    theme_animacy:recipient_animacy + 
                                    theme_pronominality:recipient_pronominality +
                                    theme_animacy:theme_pronominality +
                                    recipient_animacy:recipient_pronominality +
                                    theme_definiteness + recipient_definiteness +
                                    theme_length  + recipient_length +
                                    (1|model), data = do_pp)

summary(fit_tara_tprp_interaction_dopp, correlation = FALSE)

```

<!-- Both interactions are significant, so we can now test for main-effects individually, to test for main effect of `X` we simply fit another model with all else the same, but remove only the main effect of `X` and run a likelihood ratio test between the more complex model and this simpler model, if the more complex model is better, then that suggests `X` indeed has a main effect. -->


<!-- ### Fits for main effects -->
<!-- ```{r} -->
<!-- fit_tara_tprp_interaction_ta <- lmer(best ~ recipient_animacy + theme_animacy:recipient_animacy +  -->
<!--                                     theme_pronominality * recipient_pronominality + -->
<!--                                     theme_definiteness + recipient_definiteness + -->
<!--                                     theme_length  + recipient_length + distinctiveness + -->
<!--                                     (1|model) + (1|item_id), data = do_pp, REML = FALSE) -->

<!-- fit_tara_tprp_interaction_ra <- lmer(best ~ theme_animacy + theme_animacy:recipient_animacy +  -->
<!--                                     theme_pronominality * recipient_pronominality + -->
<!--                                     theme_definiteness + recipient_definiteness + -->
<!--                                     theme_length  + recipient_length + distinctiveness + -->
<!--                                     (1|model) + (1|item_id), data = do_pp, REML = FALSE) -->

<!-- fit_tara_tprp_interaction_tp <- lmer(best ~ theme_animacy * recipient_animacy +  -->
<!--                                     recipient_pronominality + theme_pronominality:recipient_pronominality + -->
<!--                                     theme_definiteness + recipient_definiteness + -->
<!--                                     theme_length  + recipient_length + distinctiveness + -->
<!--                                     (1|model) + (1|item_id), data = do_pp, REML = FALSE) -->


<!-- fit_tara_tprp_interaction_rp <- lmer(best ~ theme_animacy * recipient_animacy +  -->
<!--                                     theme_pronominality + theme_pronominality:recipient_pronominality + -->
<!--                                     theme_definiteness + recipient_definiteness + -->
<!--                                     theme_length  + recipient_length + distinctiveness + -->
<!--                                     (1|model) + (1|item_id), data = do_pp, REML = FALSE) -->

<!-- ``` -->

<!-- ### LRTs for main effects using ANOVAs: -->

<!-- #### Theme Animacy: -->
<!-- No evidence for effect of theme animacy ($p = .2337$). -->
<!-- ```{r} -->
<!-- anova(fit_tara_tprp_interaction, fit_tara_tprp_interaction_ta) -->
<!-- ``` -->

<!-- #### Recipient Animacy -->
<!-- Main effect of recipient animacy ($p < .001$) -->

<!-- ```{r} -->
<!-- anova(fit_tara_tprp_interaction, fit_tara_tprp_interaction_ra) -->
<!-- ``` -->


<!-- #### Theme pronominality -->
<!-- No evidence for effect of theme pronominality ($p = .9269$) -->
<!-- ```{r} -->
<!-- anova(fit_tara_tprp_interaction, fit_tara_tprp_interaction_tp) -->
<!-- ``` -->

<!-- #### Recipient Pronominality -->
<!-- Main effect of recipient prenominality ($p < .001$) -->
<!-- ```{r} -->
<!-- anova(fit_tara_tprp_interaction, fit_tara_tprp_interaction_rp) -->
<!-- ``` -->

<!-- ### Intermediate Conclusions -->
<!-- The current set of results suggest that for DO -> PP, it is largely the features of the first post-verbal argument that matter for the LMs to generalize to the alternate form. For instance, we also only find a main effect of recipient definiteness ($p < .01$) and find no evidence for its theme counterpart ($p = .2538$). The only feature for which we find main effects across both arguments is length (short is good, $p < .005$). -->

<!-- **Side Note:** It seems like the lmertest package does the LRT test for me so there was no need to fit all those separate models to get p values... (this is not possible with vanilla lme4). -->


## PP --> DO Analysis

```{r}
pp_do <- coded_results %>%
  filter(adaptation_dative == "pp", generalization_dative == "do")

fit_tara_tprp_interaction_ppdo <- lmer(best ~ 
                                    theme_animacy + theme_pronominality +
                                    recipient_animacy + recipient_pronominality +
                                    theme_animacy:recipient_animacy + 
                                    theme_pronominality:recipient_pronominality +
                                    theme_animacy:theme_pronominality +
                                    recipient_animacy:recipient_pronominality +
                                    theme_definiteness + recipient_definiteness +
                                    theme_length  + recipient_length +
                                    (1|model), data = pp_do)

summary(fit_tara_tprp_interaction_ppdo, correlation = FALSE)
```

## Attempt at displaying DO -> PP and PP -> DO in same table:

```{r}
tidy(fit_tara_tprp_interaction_dopp) %>%
  filter(effect == "fixed") %>%
  select(term, do_pp_estimate = estimate, do_pp_pval = p.value) %>%
  inner_join(
    tidy(fit_tara_tprp_interaction_ppdo) %>%
      filter(effect == "fixed") %>%
      select(term, pp_do_estimate = estimate, pp_do_pval = p.value)
  ) %>%
  mutate(
    # do_pp_estimate = format(do_pp_estimate , scientific=TRUE),
    # pp_do_estimate = format(pp_do_estimate, scientific=TRUE),
    do_pp_estimate = round(do_pp_estimate, 4),
    pp_do_estimate = round(pp_do_estimate, 4),
    do_pp_signif = case_when(
      do_pp_pval < 0.05 ~ "**",
      TRUE ~ as.character(round(do_pp_pval, 4))
    ),
    pp_do_signif = case_when(
      pp_do_pval < 0.05 ~ "**",
      TRUE ~ as.character(round(pp_do_pval, 4))
    )
  ) %>%
  select(term, do_pp_estimate, do_pp_signif, pp_do_estimate, pp_do_signif) %>%
  DT::datatable(options = list(pageLength = 14))
```


